---
title: "Recession Prediction Using Machine Learning"
author: "Monet Knupp"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: false
    theme: simplex
    df_print: paged
  pdf_document:
    toc: true
    number_sections: false
    css: styles.css
fontsize: 10pt
geometry: margin=1in
---
<style>
p {text-align: justify;} h1.title { text-align: font-weight: bold; } 
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 9,
  fig.height = 5
)
```

```{r, include=FALSE}
#==================================================================================
#Libraries and Tools

#To support data collection, preprocessing, modeling, and evaluation, I use several R packages commonly applied in financial and economic analysis. The tidyverse (including dplyr, tidyr, and ggplot2) is used for data manipulation and visualization, providing a consistent and readable workflow. The quantmod package is used to retrieve financial market data such as ETF prices, while lubridate and zoo support time series handling and date transformations. For scaling and improving visualization readability, the scales package is applied. Machine learning modeling and validation are conducted using caret, along with specific algorithms such as randomForest and glmnet for predictive modeling. Model performance is evaluated using pROC and yardstick, which provide metrics such as AUC, accuracy, and precision. Finally, knitr and kableExtra are used to generate and format the final report outputs, while broom is used to convert model results into tidy formats for interpretation. These tools collectively provide a reproducible and structured workflow for building and evaluating recession prediction models.
#===================================================================================
```

```{r }

pkgs <- c(
  "tidyverse",
  "quantmod",
  "lubridate",
  "zoo",
  "scales",
  "caret",
  "randomForest",
  "glmnet",
  "pROC",
  "yardstick",
  "knitr",
  "kableExtra",
  "broom"
)

new_pkgs <- pkgs[!pkgs %in% installed.packages()[, "Package"]]

if (length(new_pkgs) > 0) {
  install.packages(new_pkgs, repos = "https://cloud.r-project.org")
}

# Data manipulation & visualization
library(tidyverse)
library(scales)

# Time series handling
library(lubridate)
library(zoo)

# Financial data
library(quantmod)

# Machine learning
library(caret)
library(randomForest)
library(glmnet)

# Model evaluation
library(pROC)
library(yardstick)

# Reporting
library(knitr)
library(kableExtra)
library(broom)
set.seed(6574)

```


# I. Introduction

Understanding and predicting economic recessions remains a central objective in financial economics, given their significant impact on financial markets, corporate performance, investment decisions, and policy responses. Recessions influence portfolio allocation, credit risk, and business planning, making early detection especially valuable for both practitioners and policymakers. Traditionally, recession forecasting has relied on econometric models using a relatively limited set of macroeconomic indicators, such as the yield curve spread, unemployment rates, and inflation. While these models provide useful insights, they may struggle to capture more complex, nonlinear relationships across a broader set of financial and economic variables.

The central research question of this project is:

**Can machine learning classification models improve the early detection of U.S. recessions relative to a traditional econometric model using macroeconomic and financial indicators?**

To address this question, the project compares a baseline logistic regression model with machine learning approaches, including Random Forest and LASSO logistic regression. The goal is not only to evaluate statistical performance, but also to assess whether machine learning provides meaningful improvements as an early-warning system in a real-world financial context.

Building on prior feedback, the analysis expands beyond a narrow set of predictors to incorporate a richer feature space. In addition to standard macroeconomic variables (e.g., yield spreads, unemployment, and inflation), the dataset includes financial market indicators such as equity indices and other relevant signals. This broader set of predictors is particularly well-suited for machine learning methods, which are designed to handle high-dimensional data and identify the most informative variables.

From a methodological perspective, model performance is evaluated using both cross-validation techniques and an out-of-time validation framework. This approach ensures that results reflect realistic forecasting conditions rather than purely in-sample accuracy. Multiple evaluation metrics are also used to provide a more comprehensive assessment of predictive effectiveness.

From an artificial intelligence perspective, the model can be viewed as a decision-support system. It takes macro-financial data as inputs, learns patterns from historical observations, and produces a probability-based classification of recession risk. In this sense, the model aligns with the concept of an intelligent agent, where economic indicators serve as inputs from the environment and the model output supports financial decision-making.

Recession periods are defined using the [NBER-based recession indicator (USREC)](https://fred.stlouisfed.org/series/USREC?utm_source=chatgpt.com) obtained from the Federal Reserve Economic Data Federal Reserve Bank of St. Louis ([FRED](https://fred.stlouisfed.org/)) database, which serves as the dependent variable in the classification models.

```{r cars}
summary(cars)
```

## Including Plots
